---
title: "Extended Dataset Analysis"
output: html_document
date: "2025-02-06"
---

# Loading the packages and the data 

```{r}
library(dplyr)
library(ggplot2)
library(readr)

df = read.csv("originals/GH_5000.csv")

```

# Repeat the analysis performed on the initial dataset 

Change column names, turn the Haemolysed column to boolean, Date of collection as Date 
```{r}
head(df)

df = df%>% 
  rename(date_of_collection = Date.Of.Collection,
         Sex = M.F,
         Age = Age.at.test,
         IGF1_mean = IGF.I.Mean,
         P3NP_mean = PIIINP.Mean,
         GH2000_score = GH.Score,
         Athlete_Id = Athlete.ID,
         Haemolysed = Notes.re..sample.validity,
         Sample_code = Sample.Code) %>%
  mutate(X = NULL,
         Haemolysed = ifelse(Haemolysed=="Haemolysed",TRUE,FALSE), #As boolean
         date_of_collection = as.Date(date_of_collection,"%d/%m/%y")) #As date

df_multiple_tests <- df %>%
  group_by(Athlete_Id) %>%  # Group by 'new_Id'
  filter(n() > 1) %>%   # Keep only groups with more than one row
  ungroup()             # Ungroup the data

df_summary <- df_multiple_tests %>%
  group_by(Athlete_Id) %>%  # Group by 'new_Id'
  reframe(
    min_gh2000 = min(GH2000_score, na.rm = TRUE),   # Minimum of gh2000_score
    max_gh2000 = max(GH2000_score, na.rm = TRUE),   # Maximum of gh2000_score
    variance = var(GH2000_score, na.rm = TRUE),
    mean_gh2000 = mean(GH2000_score, na.rm = TRUE), # Mean of GH2000_score
    sd_gh2000 = sd(GH2000_score, na.rm = TRUE),     # Standard deviation of GH2000_score
    cutoff_point = mean(GH2000_score, na.rm = TRUE) + 3 * sd(GH2000_score, na.rm = TRUE),# Variance of gh2000_score
    num_of_repeats = n(),
    Sex = Sex,
    CV = sd(GH2000_score, na.rm = TRUE)*100/mean(GH2000_score, na.rm = TRUE)# Count of repeats
  ) %>%
  ungroup() 
```

Create a function to analyze GH2000 without the outliers

```{r}
analyze_gh2000 <- function(data, athlete_id) {  
  # Filter data for the given athlete_id
  df <- data %>%
    filter(Athlete_Id == athlete_id)  
  
  # Convert date column to Date format and sort by date
  df <- df %>% arrange(date_of_collection)
  
  # Remove outliers based on Z-score
  no_outlier <- df %>%
    mutate(z_scores = as.numeric(scale(GH2000_score))) %>%
    filter(abs(z_scores) < 2)
  
  # Compute lower and upper cutoffs (mean ± 3*SD)
  lower_cutoff <- mean(no_outlier$GH2000_score, na.rm = TRUE) - 3 * sd(no_outlier$GH2000_score, na.rm = TRUE)
  upper_cutoff <- mean(no_outlier$GH2000_score, na.rm = TRUE) + 3 * sd(no_outlier$GH2000_score, na.rm = TRUE)
  
  # Find max GH2000_score per athlete
  max_gh2000 <- max(df$GH2000_score, na.rm = TRUE)
  
  # Identify IDs exceeding the upper cutoff
  exceed_ids <- ifelse(max_gh2000 > upper_cutoff, athlete_id, NA) %>% na.omit()
  
  # Generate the plot
  plot <- ggplot(df, aes(x = date_of_collection, y = GH2000_score)) +
    geom_point(size = 3, color = "black") +                 
    geom_line(color = "black") +                           
    geom_hline(yintercept = lower_cutoff, linetype = "dotted", color = "indianred") +
    geom_hline(yintercept = upper_cutoff, linetype = "dotted", color = "indianred") +
    geom_hline(yintercept = 9.98, linetype = "solid", color = "grey") +
    geom_hline(yintercept = 8.62, linetype = "solid", color = "grey") +  
    labs(
      title = paste(athlete_id, "GH-2000 vs Date of Collection"),
      x = "Date of Collection",
      y = "GH-2000"
    ) +
    theme_minimal()
  
  # Return both the plot and the IDs that exceeded the cutoff
  return(list(plot = plot, exceed_ids = exceed_ids))
}

analyze_P3NP = function(data, athlete_id) {  
  # Filter data for the given athlete_id
  df <- data %>%
    filter(Athlete_Id == athlete_id)  
  
  # Convert date column to Date format and sort by date
  df <- df %>% arrange(date_of_collection)
  
  # Remove outliers based on Z-score
  no_outlier <- df %>%
    mutate(z_scores = as.numeric(scale(P3NP_mean))) %>%
    filter(abs(z_scores) < 2)
  
  # Compute lower and upper cutoffs (mean ± 3*SD)
  lower_cutoff <- mean(no_outlier$P3NP_mean, na.rm = TRUE) - 3 * sd(no_outlier$P3NP_mean, na.rm = TRUE)
  upper_cutoff <- mean(no_outlier$P3NP_mean, na.rm = TRUE) + 3 * sd(no_outlier$P3NP_mean, na.rm = TRUE)
  
  # Find max GH2000_score per athlete
  max_P3NP <- max(df$P3NP_mean, na.rm = TRUE)
  
  # Identify IDs exceeding the upper cutoff
  exceed_ids <- ifelse(max_P3NP > upper_cutoff, athlete_id, NA) %>% na.omit()
  
  # Generate the plot
  plot <- ggplot(df, aes(x = date_of_collection, y = P3NP_mean)) +
    geom_point(size = 3, color = "black") +                 
    geom_line(color = "black") +                           
    geom_hline(yintercept = lower_cutoff, linetype = "dotted", color = "indianred") +
    geom_hline(yintercept = upper_cutoff, linetype = "dotted", color = "indianred") +
   # geom_hline(yintercept = 9.98, linetype = "solid", color = "grey") +
   # geom_hline(yintercept = 8.62, linetype = "solid", color = "grey") +  
    labs(
      title = paste(athlete_id, "P3NP vs Date of Collection"),
      x = "Date of Collection",
      y = "P3NP Mean"
    ) +
    theme_minimal()
  
  # Return both the plot and the IDs that exceeded the cutoff
  return(list(plot = plot, exceed_ids = exceed_ids))
}

analyze_IGF1 = function(data, athlete_id) {  
  # Filter data for the given athlete_id
  df <- data %>%
    filter(Athlete_Id == athlete_id)  
  
  # Convert date column to Date format and sort by date
  df <- df %>% arrange(date_of_collection)
  
  sex = df$Sex  
  # Remove outliers based on Z-score
  no_outlier <- df %>%
    mutate(z_scores = as.numeric(scale(IGF1_mean))) %>%
    filter(abs(z_scores) < 2)
  
  # Compute lower and upper cutoffs (mean ± 3*SD)
  lower_cutoff <- mean(no_outlier$IGF1_mean, na.rm = TRUE) - 3 * sd(no_outlier$IGF1_mean, na.rm = TRUE)
  upper_cutoff <- mean(no_outlier$IGF1_mean, na.rm = TRUE) + 3 * sd(no_outlier$IGF1_mean, na.rm = TRUE)
  
  # Find max GH2000_score per athlete
  max_IGF1 <- max(df$IGF1_mean, na.rm = TRUE)
  
  # Identify IDs exceeding the upper cutoff
  exceed_ids <- ifelse(max_IGF1 > upper_cutoff, athlete_id, NA) %>% na.omit()
  
  # Generate the plot
  plot <- ggplot(df, aes(x = date_of_collection, y = IGF1_mean, colour = Haemolysed)) +
    geom_point(size = 3, color = "black") +                 
    geom_line(color = "black") +                           
    geom_hline(yintercept = lower_cutoff, linetype = "dotted", color = "indianred") +
    geom_hline(yintercept = upper_cutoff, linetype = "dotted", color = "indianred") +
    #geom_hline(yintercept = 9.98, linetype = "solid", color = "grey") +
    #geom_hline(yintercept = 8.62, linetype = "solid", color = "grey") +  
    scale_color_manual(values = c("TRUE" = "red", "FALSE" = "black")) +  # Color red if Haemolysed, black otherwise
    labs(
      title = paste(athlete_id, "- Sex:", sex, "- IGF1 vs Date of Collection"),
      x = "Date of Collection",
      y = "IGF1 Mean",
      color = "Haemolysed"  # Legend title
    ) +
    theme_minimal()
  
  # Return both the plot and the IDs that exceeded the cutoff
  return(list(plot = plot, exceed_ids = exceed_ids))
}
```






Identify Cases of high CV with multiple repeats 

```{r}
potential_cases = df_summary %>% 
  filter(
    (Sex == 'M' & CV > 15) | (Sex == 'F' & CV > 50),
    num_of_repeats > 2
  )
unique_Ids = unique(potential_cases$Athlete_Id)
unique_Ids
```

Plot the output for each suspicious athlete 

```{r}
plots <- lapply(unique_Ids, function(id) analyze_gh2000(df, id))


# Apply function across all athletes
exceeding_ids_gh <- unlist(lapply(unique_Ids, function(id) analyze_gh2000(df, id)$exceed_ids))

exceeding_ids_p3np = unlist(lapply(unique_Ids, function(id) analyze_P3NP(df, id)$exceed_ids))

exceeding_ids_igf1 = unlist(lapply(unique_Ids, function(id) analyze_IGF1(df,id)$exceed_ids))

print(exceeding_ids_p3np)  # Shows all athletes where max_gh2000 > upper_cutoff

# To display a specific plot, for example, the first one:
print(plots[13])

analyze_IGF1(df,"UKAD-2025-1261")
exceeding_ids
```

```{r}
analyze_gh2000(df, "UKAD-2025-1749")
```

Perform the Same with IGF, P3NP

Do Haemolytic samples cause outliers? (calculate percentage of values with Z>2 that are haemolytic vs % of all values that are haemolytic)

```{r}
write_csv(df,"GH_5000_Preprocessed.csv")
```


```{r}
ggplot(df, aes(x = GH2000_score, fill = as.factor(Haemolysed))) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  facet_wrap(~Haemolysed) +
  theme_minimal() +
  labs(title = "Histogram of GH2000 Score by Haemolysed Status", fill = "Haemolysed")

```

```{r}
haemolysed = df %>% filter(Haemolysed == T)


ggplot(haemolysed, aes(x = GH2000_score)) +
  geom_histogram(alpha = 0.5, position = "identity", bins = 30) +
  facet_wrap(~Haemolysed) +
  theme_minimal() 
```
```{r}
# Run Wilcoxon rank-sum test (Mann-Whitney U test)
wilcox.test(GH2000_score ~ Haemolysed, data = df)

```

```{r}
library(rcompanion)
wilcoxR(x = df$GH2000_score, g = df$Haemolysed)

```

